Binning Strategies and Related Loss for Large Data

### 2015-04-15 Paper.Rnw
Attempt to clarify notation and wording for "origin offsets".  I think the problem was univariate language for a bivariate idea. Changes in red (blue comments not deleted, just hidden)

\begin{table}[H]
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Original Data, 12 rows]{
\begin{tabular}{rr}\hline
       $x$ &  $y$  \\ \hline
-7.7325 & -9.6340 \\ 
  -8.1176 & -1.4529 \\ 
  -5.8996 & -3.2033 \\ 
  -7.0375 & -5.5563 \\ 
  -3.6354 & -3.9315 \\ 
  -8.7639 & 0.9874 \\ 
  -2.9781 & 8.6802 \\ 
  0.8210 & -8.6118 \\ 
  5.4477 & -8.4555 \\ 
  4.6849 & -5.6620 \\ 
  9.4785 & 1.1133 \\ 
  1.7579 & 5.3759 \\  \hline
\end{tabular}}
\end{minipage}\hfil
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Centers, 12 rows]{
\begin{tabularx}{.7\textwidth}{rr} \hline
  $b_X(x)$ & \hspace{.4cm} $b_Y(y)$  \\ \hline
  -5 & -5 \\ 
  -5 & -5 \\ 
  -5 & -5 \\ 
  -5 & -5 \\ 
  -5 & -5 \\ 
  -5 & 5 \\ 
  -5 & 5 \\ 
  5 & -5 \\ 
  5 & -5 \\ 
  5 & -5 \\ 
  5 & 5 \\ 
  5 & 5 \\  \hline
\end{tabularx}}
\end{minipage}\hfil
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Reduced Binned Data, 4 rows]{
\begin{tabularx}{.7\textwidth}{rrr}\hline
 \hspace{.2cm} $x^\ast$   &   \hspace{.2cm} $y^\ast$ & \hspace{.4cm}c \\ \hline
-5 & -5 & 5 \\ 
  -5 & 5 & 2 \\ 
  5 & -5 & 3 \\ 
  5 & 5 & 2 \\  \hline
\end{tabularx}}
\end{minipage}
\caption{\label{DataReductionExample}Original, Binned and Reduced Binned Data Tables, with data storage sizes.  Binned using standard rectangular approach with origin at (-10,-10) and binwidths $\omega_x = \omega_y = 10$. }
\end{table}

@@

The \textit{data resolution} is defined as the smallest increment between successive data values. To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units. This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution to which data is recorded.

@@

To demonstrate the importance of properly selecting binwidths consider the coarse version of the simulated uniform data which is recorded to a resolution of 2 units. Figure~\ref{fig:UniformStripes} displays the binned scatterplots under several scenarios. Under standard binning we see the white-space gaps with 1 X 1 unit bins, vertical and horizontal stripes with 5 X 5 unit bins and the appropriate view of an evenly spread density with 4X4 unit bins.  Note that random binning is effective at smoothing out the stripped patterns when non-integer multiples of the data resolution are selected.

@@

The binning origin can also influence the spatial loss in the binned scatterplot.  For data with fine resolution compared to the bin dimensions, the origin is only largely consequential for distributions with high density in outermost bins.  Figure~\ref{fig:ExponentialOriginPlots} displays the binned scatterplots of 10 by 10 unit bins for the fine exponential data with two different origin choices at (0,0) and (-9,-9). The binning beginning at the (-9,-9) origin suffers visually from the illusion that the density drops off near the lower bounds due to the fact that these lowest bins in each dimension are largely empty due to the negative regions. The binning origin at (0,0) is also superior in terms of the net spatial loss, which is about 7\% lower than for the (-9,-9) origin. For data with very fine resolution, it is a good idea to set the origin at the minimum value for each dimension.

@@

For data with a coarse resolution the location of the origin is also important because the origin controls the proximity of possible data values to bin centers. It can be shown analytically that for data with a resolution of $\alpha$ units with a  symmetric distribution within each of the bins, the origin that minimizes net spatial loss is located $\alpha/2$ units below the minimum value (see proof in Appendix~\ref{proof:offset}).

\hh{XXX which bin size are you talking about? add this, otherwise the statement is wrong XXX}

\hh{`$\alpha/2$ origin offset' this an implicit naming. You have to make it either explicit or write this out. At the moment it is imprecise, because you mean this offset to be below the minimal value of the data. 
XXX }

In practice the $\alpha/2$ origin offset is found to be a reasonable choice for minimizing spatial loss for symmetric bivariate data at resolution $\alpha$ while using bin dimensions that are integer multiples of $\alpha$.
\hh{In the next paragraph you are refering to the minimal value (as in minimal data value) and you are minimizing spatial loss. it's a bit confusing with the two minima floating around. It might be better to properly define the offset, and then refer to offset values rather than origins below the minimal value}
Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin is slide from the minimum value to one binwidth below the minimum value using the coarse simulated data with a resolution of 2 units. \ktm{The round glyphs indicade the origin offset where the net loss reached an absolute minimum. For the two symmetrically distributed data sets, normal and uniform, the origin offset of 1 unit was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within a 0.2\% increase from the minimum spatial loss) for each considered bin size.} %\hh{XXX - no! the value at -1 is close, but the dots are in the actual minimal values. This is only a single example - the expected minimum might still be in -1. }
For the bivariately skewed exponential data, the 1 unit origin offset minimized net loss for the smaller intervals but was not optimal for the largest intervals; 2.5\% and 7\% increase the minimum spatial loss for bins of size eight and ten, respectively.

>>>>

\begin{table}[H]
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Original Data, 12 rows]{
\begin{tabular}{rr}\hline
       $x$ &  $y$  \\ \hline
-7.7325 & -9.6340 \\ 
  -8.1176 & -1.4529 \\ 
  -5.8996 & -3.2033 \\ 
  -7.0375 & -5.5563 \\ 
  -3.6354 & -3.9315 \\ 
  -8.7639 & 0.9874 \\ 
  -2.9781 & 8.6802 \\ 
  0.8210 & -8.6118 \\ 
  5.4477 & -8.4555 \\ 
  4.6849 & -5.6620 \\ 
  9.4785 & 1.1133 \\ 
  1.7579 & 5.3759 \\  \hline
\end{tabular}}
\end{minipage}\hfil
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Centers, 12 rows]{
\begin{tabularx}{.7\textwidth}{rr} \hline
  $b_X(x)$ & \hspace{.4cm} $b_Y(y)$  \\ \hline
  -5 & -5 \\ 
  -5 & -5 \\ 
  -5 & -5 \\ 
  -5 & -5 \\ 
  -5 & -5 \\ 
  -5 & 5 \\ 
  -5 & 5 \\ 
  5 & -5 \\ 
  5 & -5 \\ 
  5 & -5 \\ 
  5 & 5 \\ 
  5 & 5 \\  \hline
\end{tabularx}}
\end{minipage}\hfil
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Reduced Binned Data, 4 rows]{
\begin{tabularx}{.7\textwidth}{rrr}\hline
 \hspace{.2cm} $x^\ast$   &   \hspace{.2cm} $y^\ast$ & \hspace{.4cm}c \\ \hline
-5 & -5 & 5 \\ 
  -5 & 5 & 2 \\ 
  5 & -5 & 3 \\ 
  5 & 5 & 2 \\  \hline
\end{tabularx}}
\end{minipage}
\caption{\label{DataReductionExample}Original, Binned and Reduced Binned Data Tables, with data storage sizes.  Binned using standard rectangular approach with origin $(\beta_{0,x},\beta_{0,y})$ = (-10,-10) and binwidths $\omega_x = \omega_y = 10$. }
\end{table}

@@

The \textit{data resolution} is defined as the smallest increment between successive data values. To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units in each dimension. This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution to which data is recorded.

@@

To demonstrate the importance of properly selecting binwidths consider the coarse version of the simulated uniform data which is recorded to a resolution of two units in each dimension. Figure~\ref{fig:UniformStripes} displays the binned scatterplots under several scenarios. Under standard binning we see the white-space gaps with one-by-one unit bins, vertical and horizontal stripes with five-by-five unit bins and the appropriate view of an evenly spread density with four-by-four unit bins.  Note that random binning is effective at smoothing out the stripped patterns when non-integer multiples of the data resolution are selected.

@@

The binning origin can also influence the spatial loss in the binned scatterplot.  For data with fine resolution compared to the bin dimensions, the origin is only largely consequential for distributions with high density in outermost bins.  Figure~\ref{fig:ExponentialOriginPlots} displays the binned scatterplots of ten-by-ten unit bins for the fine exponential data with two different origin choices at (0,0) and (-9,-9). The binning beginning at the (-9,-9) origin suffers visually from the illusion that the density drops off near the lower bounds due to the fact that these lowest bins in each dimension are largely empty due to the negative regions. The binning origin at (0,0) is also superior in terms of the net spatial loss, which is about seven percent lower than for the (-9,-9) origin. For data with very fine resolution, it is a good idea to set the origin at the minimum value for each dimension.

@@

For data with a coarse resolution the location of the origin is also important because the origin controls the proximity of possible data values to bin centers. \ktm{We define the \textit{origin offset} for each dimension, ($o_x$, $o_y$), such that 
%
\begin{eqnarray}\label{OriginOffset}
(\beta_{0,x},\beta_{0,y}) = (x_{(1)}, y_{(1)}) - (o_x, o_y),
\end{eqnarray}  
%
where $(\beta_{0,x},\beta_{0,y})$ is the bivariate bin origin, $x_{(1)}$ and $y_{(1)}$ are the minimum data value for each dimension, and $(o_x, o_y)$ is the origin offset. Thus the offset indicates the number of units in each dimension to shift the origin below the highest possible origin. }

\ktm{It can be shown analytically that an origin offset of ($\alpha_x/2$,$\alpha_y/2$) units will minimize the net spatial loss in the situation with the following three properties: (i) data are recorded to a resolution of $\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension, (ii) points are symmetric distributed within rectangular bins, (iii) the bin dimesions are integer multiples of $\alpha_x$ and $\alpha_y$, respectively (see proof in Appendix~\ref{proof:offset}). }

%\hh{`$\alpha/2$ origin offset' this an implicit naming. You have to make it either explicit or write this out. At the moment it is imprecise, because you mean this offset to be below the minimal value of the data. XXX }

In practice the ($\alpha_x/2$,$\alpha_y/2$) origin offset is found to be a reasonable choice for lowering spatial loss for symmetric bivariate data at a coarse resolution while using bin dimensions that are integer multiples of $\alpha_x$ and $\alpha_y$.
%\hh{In the next paragraph you are refering to the minimal value (as in minimal data value) and you are minimizing spatial loss. it's a bit confusing with the two minima floating around. It might be better to properly define the offset, and then refer to offset values rather than origins below the minimal value}
\ktm{Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin offset is changed while binning the coarse simulated data using the standard rectangular binning algorithm. Note that for simplicity, only origin offsets for with equal offset in each dimension are included in this simulation. Since the data resolution is two units in each dimension, we will pay special attention to an origin offset of (1,1) in order to assess how well the theoretically optimal origin offset at ($\alpha_x/2$,$\alpha_y/2$) works for data that violate the assumptions to differing degrees. The round glyphs indicade the origin offset where the net loss reached an absolute minimum. For the two symmetrically distributed data sets, normal and uniform, the origin offset of one unit was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within a 0.2\% increase from the minimum spatial loss) for each considered bin size.} %\hh{XXX - no! the value at -1 is close, but the dots are in the actual minimal values. This is only a single example - the expected minimum might still be in -1. }
For the bivariately skewed exponential data, the 1 unit origin offset minimized net loss for the smaller intervals but was not optimal for the largest intervals; 2.5\% and 7\% increase the minimum spatial loss for bins of size eight and ten, respectively.
